---
title: "Tomales Bay Winter Shorebird annual summary"
author: "CGRC Staff - Scott Jennings"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      error = FALSE,
                      warning = FALSE)
```


```{r}

# packages, source ----
library(tidyverse)
library(lubridate)
library(MASS)
library(here)

library(birdnames)
custom_bird_list <- readRDS("C:/Users/scott.jennings/OneDrive - Audubon Canyon Ranch/Projects/my_R_general/birdnames_support/data/custom_bird_list")

# local copy of this function https://github.com/ateucher/useful_code/blob/master/R/numbers2words.r
source("C:/Users/scott.jennings/OneDrive - Audubon Canyon Ranch/Projects/my_R_general/utility_functions/numbers2words.R")

options(scipen = 999)

zyear = 2022

delta_year = numbers2words(zyear - 2019)

exclude_date <- as.Date(c("1990-01-04", "1990-02-12", "2010-01-18"))
```

```{r}
sbirds <- readRDS(here("data_files/rds/shorebirds_for_analysis")) %>% 
  rename(section = North_South_Code)

rapt <- readRDS(here("data_files/rds/raptors4analysis")) %>%     
  mutate(date = as.Date(date)) %>% 
  rename(section = North_South_Code) %>% 
  filter(!(date %in% exclude_date)) %>% 
  filter(grepl("winter", season.year), bird.hunter == 1, count >= 0) %>% 
  #filter(grepl("winter", season.year), species %in% c("PEFA", "MERL"), count >= 0, season.year != "winter_2019") %>% 
  group_by(date, section) %>% 
  summarise(tot.count = sum(count)) %>% 
  ungroup() %>% 
  mutate(year = ifelse(month(date) < 6, year(date)-1, year(date))) %>% 
  group_by(year, section) %>% 
  summarise(mean.bird.hunters = mean(tot.count),
            sd.bird.hunters = sd(tot.count)) %>% 
  ungroup() 

# rain from C:/Users/scott.jennings/OneDrive - Audubon Canyon Ranch/Projects/general_data_sources/Rainfall/get_PRISM.R
rain <- read.csv("C:/Users/scott.jennings/OneDrive - Audubon Canyon Ranch/Projects/general_data_sources/Rainfall/derived_data/tomales_mean_month_rain.csv") %>% 
  mutate(year = ifelse(month < 7, year - 1, year)) %>% 
  filter(month <= 2 | month > 7) %>% 
  group_by(year) %>% 
  summarise(seas.rain.mm = sum(mean.rain),
            seas.rain.mm = round(seas.rain.mm)) %>% 
  filter(year > 1988)


# annual estimate each species
analysis_table <- sbirds %>% 
  group_by(season.year, date, alpha.code) %>% 
  summarize(total.sbirds = sum(count)) %>% 
  ungroup() %>% 
  group_by(season.year, alpha.code) %>% 
  summarise(mean.total.sbirds = floor(mean(total.sbirds)),
            sd.total.sbirds = sd(total.sbirds),
            max.total.sbirds = max(total.sbirds),
            p75.total.sbirds = quantile(total.sbirds, probs = c(0.75)),
            num.surveys = n()) %>% 
  ungroup() %>%
  bind_rows(., sbirds %>% 
              group_by(season.year, date) %>% 
              summarize(total.sbirds = sum(count)) %>% 
              ungroup() %>% 
              group_by(season.year) %>% 
              summarise(mean.total.sbirds = floor(mean(total.sbirds)),
                        sd.total.sbirds = sd(total.sbirds),
                        max.total.sbirds = max(total.sbirds),
                        p75.total.sbirds = quantile(total.sbirds, probs = c(0.75)),
                        num.surveys = n()) %>%
              ungroup() %>% 
              mutate(alpha.code = "All")) %>% 
  separate(season.year, c("season", "year"), remove = F) %>% 
  mutate(year = as.numeric(year),
         giac.dummy = ifelse(year < 2009, 0, 1)) %>% 
  full_join(., rain)



znewdat <- analysis_table %>% 
  #distinct(year, section, giac.dummy) %>% 
  distinct(year) %>% 
  mutate(#giac.dummy = ifelse(year < 2009, 0, 1),
         seas.rain.mm = mean(rain$seas.rain.mm)) #%>% 
#  full_join(mean_preds)

```

```{r}

zmaxit = 400

fit_big_mod <- function(zspp) {
  sbirds_winter <- analysis_table %>% 
    filter(season == "winter", alpha.code == zspp)
  
# big_mod <- glm.nb(floor(p75.total.sbirds) ~ poly(year, 2) * section + giac.dummy + seas.rain.mm + mean.bird.hunters, data = sbirds_winter, maxit = zmaxit)

big_mod <- glm.nb(floor(p75.total.sbirds) ~ poly(year, 2) + seas.rain.mm, data = sbirds_winter, maxit = zmaxit)

}

# ---
sbird_predicter_glm <- function(zmod, newdat){
ilink <- family(zmod)$linkinv
 all_best_pred = predict(zmod, newdat, se.fit=TRUE, type='link') %>% 
  data.frame() %>% 
  cbind(newdat) %>% 
  ungroup() %>% 
  mutate(predicted = ilink(fit),
         lci = ilink(fit - (1.96 * se.fit)),
         uci = ilink(fit + (1.96 * se.fit)))
  }


# ---

sbird_mod_plotter <- function(zpred) {
if(zpred$alpha.code[1] == "All") {
  sp_name = "All shorebirds"
} else {
sp_name <- translate_bird_names(zpred$alpha.code[1], "alpha.code", "common.name")
}
  
zpred <- zpred %>% 
  left_join(., analysis_table %>% filter(season == "winter") %>% dplyr::select(alpha.code, year, p75.total.sbirds)) %>% 
  rename(Year = year)

ymax = ifelse(max(zpred$p75.total.sbirds) > max(zpred$uci), max(zpred$p75.total.sbirds), max(zpred$uci))

zplot <- zpred %>% 
  ggplot(aes(x = Year, y = predicted)) +
  geom_line() +
  geom_ribbon(aes(ymin=lci,ymax=uci),alpha=0.3) +
    geom_point(aes(x = Year, y = p75.total.sbirds)) +
  theme_bw() +
  scale_x_continuous(limits = c(1989, zyear), breaks = c(1990, 2000, 2010, 2020)) +
  xlab("Year") +
  ylab("Estimated mean abundance") +
  ggtitle(sp_name) +
  ylim(0, ymax)

zplot

}


fit_plot_mod <- function(zspp) {
  
 spp_mod <- fit_big_mod(zspp)

spp_pred <- sbird_predicter_glm(spp_mod, znewdat) %>% 
  mutate(alpha.code = zspp)


wide_preds <- spp_pred %>% 
  filter(year %in% c(1989, zyear)) %>% 
  mutate(predicted = round(predicted, 0),
         year = ifelse(year == zyear, "zyear", year),
         year = paste("yr.", year, sep = "")) %>% 
  pivot_wider(id_cols = alpha.code, names_from = year, values_from = predicted) %>% 
  mutate(change = round(yr.zyear, 0) - round(yr.1989, 0),
         per.change = 100*(change/round(yr.1989, 0)),
         per.change = round(per.change, 1)) %>% 
  mutate(out.spp = ifelse(zspp == "All", " shorebirds (all species combined)", translate_bird_names(zspp, "alpha.code", "common.name")),
         out.text = paste("In ", zyear, ", there were an estimated ", yr.zyear, " ", out.spp, " on Tomales Bay, compared to ", yr.1989, " in 1989 (", per.change, "% change).", sep = "")) 

sbird_out <- list(zplot = sbird_mod_plotter(spp_pred),
                  ztext = wide_preds$out.text,
                  all.pred = dplyr::select(spp_pred, year, predicted, lci, uci, alpha.code))
}

```

```{r}

zspp_list <- c("All", "DUNL", "WESA", "MAGO", "LESA", "SAND", "WILL", "DOSP", "BBPL", "BLTU", "YELL", "SEPL", "KILL", "SPSA")

spp_out <- map(zspp_list, fit_plot_mod)

names(spp_out) <- zspp_list

get_preds <- function(zspp_out) {
  zpred <- zspp_out[[3]]
}

all_preds <- map_df(spp_out, get_preds) %>% 
  group_by(year, alpha.code) %>% 
  summarise(tot.predicted = floor(sum(predicted))) %>% 
  ungroup() %>% 
  filter(year %in% c(1989, 2019, zyear)) %>% 
  mutate(year = ifelse(year == zyear, "zyear", year),
         year = paste("yr.", year, sep = "")) %>% 
  pivot_wider(id_cols = alpha.code, names_from = year, values_from = tot.predicted) %>% 
  mutate(yr.19.change = ((yr.1989 - yr.2019)/yr.1989) * 100,
         zyr.change = ((yr.1989 - yr.zyear)/yr.1989) * 100)


```


Monitoring species populations provides critical information to identify imperiled species, evaluate current or past management, prioritize new conservation measures.  

Each winter since 1989, Audubon Canyon Ranch biologists and volunteers have monitored the number of shorebirds at Tomales Bay. We survey all shorebird habitat on the bay up to 6 times each winter. From these data we can see how shorebird populations have changed in that time, and the results are sobering. In early 2021 we published a paper titled "Declining wintering shorebird populations at a temperate estuary in California: A 30 year perspective" (full citation and link below), in which we reported an overall 66% decline in Tomales Bay shorebirds. Two of the most abundant species, Western Sandpiper and Dunlin, declined by 94% and 77%, respectively.

Warnock, N., Jennings, S., Kelly, J.P., Condeso, T.E. and Lumpkin, D., 2021. Declining wintering shorebird populations at a temperate estuary in California: A 30-year perspective. The Condor, 123(1), p.duaa060. https://academic.oup.com/condor/article/123/1/duaa060/6132586

Here, we provide an update to the results of that paper, including the subsequent `r delta_year` years of data. As in the published paper, we consider All species combined and the 12 most abundant species individually. It should be noted that the numbers we report here are slightly different than those in the paper because here we modeled the whole bay together rather than the North and South sections separately, and we only consider the extra effect of rainfall on shorter term abundance fluctuations (rainfall was identified as the most important environmental predictor in our paper). Overall, the declines we reported in the paper have continued, especially for all shorebird species combined, Western Sandpiper, Sanderling, Willet, Black-bellied Plover, Black Turnstone, and Killdeer. Contrary to what we reported in the paper, with three additional years of data Marbled Godwits appear to be in decline. An optimistic difference seen here is that the decline in Dunlin abundance appears to be leveling off, with relatively stable numbers over the last 4-5 years. Also optimistically, the increases we previously reported for Least Sandpipers and Yellowlegs have continued. However, numbers of both species have seen somewhat low in recent years so the estimated trends are a little flatter than we reported three years ago.


\newpage  

`r spp_out$All$ztext`

```{r}
spp_out$All$zplot

```

\newpage  


`r spp_out$DUNL$ztext`

```{r}
spp_out$DUNL$zplot
```

\newpage  


`r spp_out$WESA$ztext`

```{r}
spp_out$WESA$zplot
```

\newpage  

`r spp_out$MAGO$ztext`

```{r}
spp_out$MAGO$zplot
```

\newpage  

`r spp_out$LESA$ztext`

```{r}
spp_out$LESA$zplot
```

\newpage  

`r spp_out$SAND$ztext`

```{r}
spp_out$SAND$zplot
```

\newpage  

`r spp_out$WILL$ztext`

```{r}
spp_out$WILL$zplot
```


\newpage  

`r spp_out$DOSP$ztext`

```{r}
spp_out$DOSP$zplot
```

\newpage  


`r spp_out$BBPL$ztext`

```{r}
spp_out$BBPL$zplot
```

\newpage  

`r spp_out$BLTU$ztext`

```{r}
spp_out$BLTU$zplot
```


\newpage  

`r spp_out$YELL$ztext`

```{r}
spp_out$YELL$zplot
```

\newpage  

`r spp_out$SEPL$ztext`

```{r}
spp_out$SEPL$zplot
```

\newpage  

`r spp_out$KILL$ztext`

```{r}
spp_out$KILL$zplot
```


